Transferability of clinical prediction models for early trauma care

### 2019-09-09 ManuscriptMarkdown.Rmd
Update ManuscriptMarkdown.Rmd

---
title: "Transferability of clinical prediction models for early trauma care"
author: 
- Martin Henriksson
- Dell D Saulnier
- Johanna Berg
- Martin Gerdin Wärnberg
date: "August 14, 2019"
documentclass: "article"
output:
  word_document:
    reference_docx: transferstyle.docx
header-includes:
   - \usepackage{hyperref}
bibliography: library.bib
csl: bmcemerg.csl
---
```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)

@@

# Discussion
This study aimed to assess how transfers of clinical prediction models for early
trauma care between different contexts within a single health system affect
mistriage rates. In many data sets, varying degrees of uncertainty were observed 
when comparing the transferred performance to the receiving sample validation performance.

The most notable effect on model performance following model transfer were observed
after transferring the `r colnames(highest.comp.mistriage)` model. This transfer
resulted in an increase in the mistriage rate of 
`r substr(format(highest.comp.mistriage), 1, 5)`. Mainly contributing to this was
an increase in overtriage.  For this transfer, the 95% UI ranged from 
`r UI.top.bot[1]` to `r UI.top.bot[2]`, indicating a high level of uncertainty. 
In contrast, the transfer of the high-volume model to the low-volume sample 
reduced mistriage, this was primarily due to reduced overtriage.

Low-volume centres do by definition receive fewer trauma patients, and therefore 
less data is available to establish accurate clinical prediction models. It is 
reasonable to assume this is the reason for the marked increase in mistriage 
when transferring the low-volume model to the high-volume sample. This also 
explains why the low-volume model and the single-centre model showed the highest 
levels of mistriage during model validation (i.e. when applied to the sample in 
which the model was created). Further confirming this, the transfer of the 
high-volume model to the low-volume sample reduces mistriage.

In a clinical context, this seems to suggest clinical prediction models for early 
trauma care should be developed in settings with a high number of trauma patients. 
These models could then be transferred to settings with less trauma patients. Low-volume 
centres generally have fewer resources and less experience receiving trauma patients, 
these centres could therefore benefit by using models developed in centres with more 
experience receiving trauma patients

Recognizing settings with high mistriage and updating models accordingly could lead
to increased accuracy during trauma patient triage, potentially saving lives by
minimizing undertriage, and likely saving resources by minimizing overtriage. With
registries such as SweTrau, the analysis of model transfer could be performed
continuously to optimize the models currently in use and advise which model
transfers might be detrimental (or potentially beneficial).

@@

External validation as the evaluation of a clinical prediction model in a
setting in which it was not originally developed has been studied more
extensively. Studies using simulated clinical prediction models based on
different predictors showed a decline in model performance when externally
validated [@RN60; @RN61]. A study from 2018 examining the external validity of
prediction models for coronary artery disease (CAD) [@RN62] showed that the models
significantly underestimated the probability of CAD. The combined results of
this study, the studies on model transfer and the studies on external validation
would suggest caution when transferring models.

## Strengths and limitations
### Strengths
The main strength of this study is that its design realistically reflects
potential model transfers: clinical prediction models for trauma are usually
developed in large metropolitan centres with a high volume of patients and 
then transferred as clinical recommendations to minor low-volume centres. The 
results should therefore be of practical importance.

>>>>

---
title: "Transferability of clinical prediction models for early trauma care"
author: 
- Martin Henriksson
- Dell D Saulnier
- Johanna Berg
- Martin Gerdin Wärnberg
date: "September 09, 2019"
documentclass: "article"
output:
  word_document:
    reference_docx: transferstyle.docx
header-includes:
   - \usepackage{hyperref}
bibliography: library.bib
csl: bmcemerg.csl
---
```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)

@@

# Discussion
This study aimed to assess how transfers of clinical prediction models for early
trauma care between different contexts within a single health system affect
mistriage rates. Varying degrees of uncertainty were observed when comparing 
the transferred performance to the receiving sample validation performance.

The most notable effect on model performance following model transfer was observed
after transferring the `r colnames(highest.comp.mistriage)` model. This transfer
resulted in an increased mistriage rate of `r substr(format(highest.comp.mistriage), 1, 4)`.
Mainly contributing to this was
an increase in overtriage.  For this transfer, the 95% UI ranged from 
`r UI.top.bot[1]` to `r UI.top.bot[2]`, indicating a high level of uncertainty. 
In contrast, the transfer of the high-volume model to the low-volume sample 
reduced mistriage, primarily due to reduced overtriage.

Low-volume centres do by definition receive fewer trauma patients, and therefore 
less data is available to establish accurate clinical prediction models. It is 
reasonable to assume this is the reason for the marked increase in mistriage 
when transferring the low-volume model to the high-volume sample. This also 
explains why the low-volume model and the single-centre model showed the highest 
levels of mistriage during model validation (i.e. when applied to the sample in 
which the model was created). Further confirming this, the transfer of the 
high-volume model to the low-volume sample reduced mistriage.

In a clinical context, this suggests that clinical prediction models for early 
trauma care should be developed in settings with a high number of trauma patients.
These models could then be transferred to settings with less trauma patients, which 
should then result in a decreased mistriage rate in those settings.

In 2018, Granström et al. found that using a criteria-based triage system based on a 
clinical prediction model was effective in reducing overtriage without increasing 
mortality [@RN44]. Therefore, improving clinical prediction models to reduce mistriage is
clinically beneficial, potentially saving lives by minimizing undertriage, and likely 
saving resources by minimizing overtriage. Possibly, with registries as SweTrau, the analysis of 
model transfer could be performed continuously to optimize the models currently in use 
and advise which model transfers might be detrimental (or potentially beneficial).

@@

External validation - the evaluation of a clinical prediction model in a
setting in which it was not originally developed - has been studied more
extensively. Studies using simulated clinical prediction models based on
different predictors showed a decline in model performance when externally
validated [@RN60; @RN61]. A study from 2018 examining the external validity of
prediction models for coronary artery disease (CAD) [@RN62] showed that the models
significantly underestimated the probability of CAD. The combined results of
this study, the studies on model transfer and the studies on external validation
would suggest indiscriminate model transfers could lead to detrimental clinical
consequences, but also that clinically beneficial model transfers could be
identified.

## Strengths and limitations
### Strengths
The study design realistically reflects
potential model transfers: clinical prediction models for trauma are usually
developed in large metropolitan centres with a high volume of patients and 
then transferred as clinical recommendations to minor low-volume centres. The 
results should therefore be of practical importance.
